High geomagnetic field intensity recorded by anorthosite xenoliths requires a vigorous late Mesoproterozoic geodynamo

### 2021-12-20 PNAS_manuscript/PNAS_manuscript.tex

In addition to estimating paleointensities by introducing a set of selection criteria to filter our experiment results, we apply an independent method of ref. \citealp{Cych2021a} which uses all experimental data regardless of their NRM/TRM plot statistics to perform bias corrected estimation of paleointensity (BiCEP). This Bayesian probabilistic method is based on an assumption that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Arai1963a, Paterson2011a}. The posterior paleointensity distributions for sites with high-quality specimen-level data (Fig. \ref{fig:PINT_cooling_corrected}) are in agreement with site-level weighted-mean results (Fig. S3).

@@

Rock magnetic data support that the anorthosite specimens that pass the paleointensity selection criteria have dominant magnetic remanence carriers with magnetic properties similar to stochiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection and all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show both diabase and anorthosite have dominant magnetic mineralogy of magnetite as evidenced through the Verwey transition (Fig. S1; ref. \citealp{Verwey1939a}). More, anorthosite chips from sites that yield successful paleointensity results have Verwey transition temperatures near 120 Kelvin (K) as expected for stochiometric low-Ti magnetite \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have suppressed Verwey transition toward lower temperatures (Fig. S1), indicating that the specimens likely have a magnetic mineralogy with variable degrees of oxidation or relatively high Ti content \cite{Ozdemir1993a}.

@@

\subsection*{Cooling rate correction}
Paleointensity estimates from igneous intrusions can also be biased toward higher values due to slow cooling rates \cite{Halgedahl1980a}. It has been shown that large differences in cooling rates between acquisition of an NRM versus a laboratory pTRM can result in overestimated paleointensities for SD grains \cite{Dodson1980a, Halgedahl1980a, Nagy2021a}. From the thermal history model of ref. \citealp{Zhang2021b} we can estimate the duration which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they fully block in their characteristic natural remanence magnetization ($\sim$500\textdegree C; ref. \citealp{Zhang2021b}). We find the cooling time to be on the order of $\sim$1 kyr, which corresponds to a cooling rate of $\sim 1.6\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster, at $\sim 1.3\times10^{-1}$ $^\circ$C s$^{-1}$, assuming that it takes 1000s in lab condition to cool through the same temperature interval. This significant cooling rate difference leads to a predicted $\sim$36\% overestimate of true ancient field following the model of ref. \citealp{Halgedahl1980a} (Fig. S2). This estimate on cooling rate effect is similar to that derived from the model of ref. \citealp{Nagy2021a}. We therefore correct our paleointensity results by a factor of 0.74. The cooling rate-corrected specimen paleointensity estimates together with specimen- and site-level weighted means are shown in Fig. \ref{fig:PINT_cooling_corrected}.

@@

The high geomagnetic field estimates from the Midcontinent Rift volcanics and intrusives are not the only suite of high paleointensity records at odds with the hypothesis of a monotonically decaying geodynamo through the entire Proterozoic Eon \cite{Biggin2009a, Bono2019a}. The $\sim$60 ZAm$^2$ virtual dipole moment estimates from the \textit{ca.} 755 Ma Mundine Well dikes are much higher than the values predicted by the monotonic decrease model of \cite{Bono2019a}. Although one may consider these data capturing relatively short term fluctuations of the geomagnetic field, they nevertheless indicate that the geodynamo had strong power sources during the Proterozoic. On the other hand, besides the low paleointensity values recorded by the \textit{ca.} 1140 Ma Abitibi dikes, low estimates have been obtained from the \textit{ca.} 1070 Ma Bangemall Sills \cite{Lloyd2021b}, the \textit{ca.} 720 Ma dolerite dikes of the Franklin Large Igneous Province \cite{Lloyd2021a}, as well as Devonian-aged rocks from the Minusa Basin, Kola Peninsula, and Buribay volcanic complex (Fig. \ref{fig:PINT_compilation}; ref. \citealp{Shcherbakova2017a, Shcherbakova2021a}). The multiple occurrences of low geomagnetic field intensities during the Proterozoic and the Paleozoic indicate that a waning thermal convection regime in Earth's before the emergence of the solid inner core may not be the solely possible mechanism for triggering an weak dipole field.

@@

\subsection*{Paleointensity experiment}
A total of 86 specimens from 14 anorthosite xenoliths and a total of 69 specimens from 7 diabase sites underwent paleointensity experiments that followed the step-wise double-heating Thellier method \cite{Thellier1959a} using the IZZI protocol \cite{Yu2004a} with heating steps up to 585 \textdegree C. Partial thermal remanent magnetization (pTRM) checks were performed systematically throughout the experiment to test whether there was significant mineralogical alteration due to heating and were assessed using the SCAT parameter of ref. \citealp{Shaar2013a}. On top of the IZZI-Thellier experiment protocol, we also performed a comparative study where we added an extra step of 20 mT alternating field (AF) cleaning on some of the specimens after each in-field step. The purpose is to study whether the AF cleaning could help improve experiment success rate by removing the remanence component carried by materials such as multi-domain (MD) grains that contribute to non-ideal paleointensity behaviors. All remanence measurements were made on a 2G Enterprises DC-SQUID superconducting rock magnetometer equipped with an automated sample changer system at the UC Berkeley Paleomagnetism laboratory. The magnetometer is housed inside a three-layer magnetostatic shield that maintains background fields of less than 500 nT. Heating steps were performed using an ASC TD-48SC thermal demagnetizer with a controlled field coil that allows for a magnetic field to be generated in the oven in conjunction with a DC power supply. The thermal demagnetizer was degaussed with an alternating field in the axial orientation following each in-field step such that residual fields within the oven were $<$10 nT during zero-field steps. Samples were placed in the same location within the thermal demagnetizer for each heating step and were maintained in the same orientation with regard to the applied field. During each heating step, the oven remained at peak temperatures for 20 min to make sure each specimen reached the target temperature. An applied laboratory field of 30 $\mu$T was used for all in-field steps. All heating steps were performed in air. The temperature increments for the experiments were chosen to isolate magnetizations held by (titano)magnetite informed by the previous demagnetization data, with smaller increments performed close to the expected unblocking temperature of $\sim$580\textdegreeC \cite{Zhang2021b}.

>>>>

In addition to estimating paleointensities by introducing a set of selection criteria to filter our experiment results, we apply an independent method of ref. \citealp{Cych2021a} which uses all experimental data regardless of their NRM/TRM plot statistics to perform bias corrected estimation of paleointensity (BiCEP). This Bayesian probabilistic method is based on an assumption that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Paterson2011a}. The posterior paleointensity distributions for sites with high-quality specimen-level data are in agreement with site-level weighted-mean results (Fig. \ref{fig:PINT_cooling_corrected}; Fig. S1).

@@

Rock magnetic data support that the anorthosite specimens that pass the paleointensity selection criteria have dominant magnetic remanence carriers with magnetic properties similar to stochiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection and all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show both diabase and anorthosite have dominant magnetic mineralogy of magnetite as evidenced through the Verwey transition (Fig. S2; ref. \citealp{Verwey1939a}). More, anorthosite chips from sites that yield successful paleointensity results have Verwey transition temperatures near 120 Kelvin (K) as expected for stochiometric low-Ti magnetite \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have suppressed Verwey transition toward lower temperatures (Fig. S2), indicating that the specimens likely have a magnetic mineralogy with variable degrees of oxidation or relatively high Ti content \cite{Ozdemir1993a}.

@@

\subsection*{Cooling rate correction}
Paleointensity estimates from igneous intrusions can also be biased toward higher values due to slow cooling rates \cite{Halgedahl1980a}. It has been shown that large differences in cooling rates between acquisition of an NRM versus a laboratory pTRM can result in overestimated paleointensities for SD grains \cite{Dodson1980a, Halgedahl1980a, Nagy2021a}. From the thermal history model of ref. \citealp{Zhang2021b} we can estimate the duration which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they fully block in their characteristic natural remanence magnetization ($\sim$500\textdegree C; ref. \citealp{Zhang2021b}). We find the cooling time to be on the order of $\sim$1 kyr, which corresponds to a cooling rate of $\sim 1.6\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster, at $\sim 1.3\times10^{-1}$ $^\circ$C s$^{-1}$, assuming that it takes 1000s in lab condition to cool through the same temperature interval. This significant cooling rate difference leads to a predicted $\sim$36\% overestimate of true ancient field following the model of ref. \citealp{Halgedahl1980a} (Fig. S3). This estimate on cooling rate effect is similar to that derived from the model of ref. \citealp{Nagy2021a}. We therefore correct our paleointensity results by a factor of 0.74. The cooling rate-corrected specimen paleointensity estimates together with specimen- and site-level weighted means are shown in Fig. \ref{fig:PINT_cooling_corrected}.

@@

The high geomagnetic field estimates from the Midcontinent Rift volcanics and intrusives are not the only suite of high paleointensity records at odds with the hypothesis of a monotonically decaying geodynamo through the entire Proterozoic Eon \cite{Biggin2009a, Bono2019a}. The $\sim$60 ZAm$^2$ virtual dipole moment estimates from the \textit{ca.} 755 Ma Mundine Well dikes are much higher than the values predicted by the monotonic decrease model of \cite{Bono2019a}. Although one may consider these data capturing relatively short term fluctuations of the geomagnetic field, they nevertheless indicate that the geodynamo had strong power sources during the Proterozoic. On the other hand, besides the low paleointensity values recorded by the \textit{ca.} 1140 Ma Abitibi dikes, low estimates have been obtained from the \textit{ca.} 1070 Ma Bangemall Sills \cite{Lloyd2021b}, the \textit{ca.} 720 Ma dolerite dikes of the Franklin Large Igneous Province \cite{Lloyd2021a}, as well as Devonian-aged rocks from the Minusa Basin, Kola Peninsula, Buribay volcanic complex and volcanic rocks from the East European Craton (Fig. \ref{fig:PINT_compilation}; ref. \citealp{Shcherbakova2017a, Shcherbakova2019a, Shcherbakova2021a}). The multiple occurrences of low geomagnetic field intensities during the Proterozoic and the Paleozoic indicate that a waning thermal convection regime in Earth's before the emergence of the solid inner core may not be the solely possible mechanism for triggering an weak dipole field.

@@

\subsection*{Paleointensity experiment}
A total of 86 specimens from 14 anorthosite xenoliths and a total of 69 specimens from 7 diabase sites underwent paleointensity experiments that followed the step-wise double-heating Thellier method \cite{Thellier1959a} using the IZZI protocol \cite{Yu2004a} with heating steps up to 585 \textdegree C. Partial thermal remanent magnetization (pTRM) checks were performed systematically throughout the experiment to test whether there was significant mineralogical alteration due to heating and were assessed using the SCAT parameter of ref. \citealp{Shaar2013a}. On top of the IZZI-Thellier experiment protocol, we also performed a comparative study where we added an extra step of 20 mT alternating field (AF) cleaning on some of the specimens after each in-field step. The purpose is to study whether the AF cleaning could help improve experiment success rate by removing the remanence component carried by materials such as multi-domain (MD) grains that contribute to non-ideal paleointensity behaviors. All remanence measurements were made on a 2G Enterprises DC-SQUID superconducting rock magnetometer equipped with an automated sample changer system at the UC Berkeley Paleomagnetism laboratory. The magnetometer is housed inside a three-layer magnetostatic shield that maintains background fields of less than 500 nT. Heating steps were performed using an ASC TD-48SC thermal demagnetizer with a controlled field coil that allows for a magnetic field to be generated in the oven in conjunction with a DC power supply. The thermal demagnetizer was degaussed with an alternating field in the axial orientation following each in-field step such that residual fields within the oven were $<$10 nT during zero-field steps. Samples were placed in the same location within the thermal demagnetizer for each heating step and were maintained in the same orientation with regard to the applied field. During each heating step, the oven remained at peak temperatures for 20 min to make sure each specimen reached the target temperature. An applied laboratory field of 30 $\mu$T was used for all in-field steps. All heating steps were performed in air. The temperature increments for the experiments were chosen to isolate magnetizations held by (titano)magnetite informed by the previous demagnetization data, with smaller increments performed close to the expected unblocking temperature of $\sim$580\textdegree C \cite{Zhang2021b}.

### 2021-12-20 PNAS_manuscript/Supporting Info.tex
update manuscript

@@

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{manuscript/Cooling_rate_correction.pdf}
\centering
\caption{{Graph of predicted paleointensity overestimate due to slow cooling of the intrusive Beaver River diabase and anorthosite xenoliths relative to the cooling rate in laboratory following the model of \citeA{Halgedahl1980a}. Because the majority of the anorthosites have unblocking temperatures between 500\textdegree C and 580\textdegree C, we estimate that the slow cooling during natural remanence acquisition could have resulted in a 36\% overestimate. Thus, a correction factor of 0.74 is applied at specimen level for the paleointensity summary plot.  }}
\label{fig:PINT_cooling_corrected}
\end{figure*}

@@

>>>>

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{manuscript/PINT_BiCEP.pdf}
\centering
\caption{{Violin plots of site-level posterior paleointensity distributions estimated using the bias corrected estimation of paleointensity (BiCEP) method developed by ref. \ref{Cych2021a}. Assuming that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Arai1963a, Paterson2011a}, the method uses all paleointensity measurement-level data without applying selection criteria. For comparison of results from this independent method with those based on our selection (as show in Fig. 4 in manuscript), we highlight the anorthosite sites that pass our paleointensity selection and make other anorthosite and diabase transparent. The results from the BiCEP method address the uncertainties associated with anorthosite AX6 and AX8. This is associated with the relatively variable specimen behaviors within these two sites. But for sites AX10, AX11, AX13, AX12, and AX16, the posterior probability distributions have very narrow bounds, consistent with the interpretation that these anorthosites are faithful paleointensity recorders that have high-quality paleointensity behaviors. }}
\label{fig:PINT_BiCEP}
\end{figure*}

@@

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{Cooling_rate_correction.pdf}
\centering
\caption{{Graph of predicted paleointensity overestimate due to slow cooling of the intrusive Beaver River diabase and anorthosite xenoliths relative to the cooling rate in laboratory following the model of \citeA{Halgedahl1980a}. Because the majority of the anorthosites have unblocking temperatures between 500\textdegree C and 580\textdegree C, we estimate that the slow cooling during natural remanence acquisition could have resulted in a 36\% overestimate. Thus, a correction factor of 0.74 is applied at specimen level for the paleointensity summary plot.  }}
\label{fig:PINT_cooling_corrected}
\end{figure*}

@@

% \begin{figure*}[h!]
% \noindent\includegraphics[width=17.8 cm]{manuscript/nonlinear.pdf}
% \centering
% \caption{{}}
% \label{fig:nonlinear_check}
% \end{figure*}

% \clearpage
